---
title: "Day 7 activities"
format: html
editor: visual
---

```{r}
library(datapasta)
library(tidyverse)
library(janitor) 
library(here)
library(palmerpenguins)
library(lubridate)
library(here)
```

# task 1

```{r}
#read in the datasets

bird_observations <- read_csv(here::here("data", "52_pp52_birds_1.csv"))

sites <- read_csv(here::here("data", "52_pp52_sites_1.csv"))

surveys <- read_csv(here::here("data", "52_pp52_surveys_1.csv"))

taxalist <- read_csv(here::here("data", "52_pp52_taxalist_1.csv"))


  
```

### left join

```{r}
#create subset for only birds with species ID: "BHCO" and "RWBL", from only sites: "LI-W" and "NU-C"

birds_subset <- bird_observations |> 
  filter(species_id %in% c("BHCO", "RWBL")) |> 
  filter(site_id %in% c("LI-W", "NU-C"))

#Use left join(s) to update birds_subset so that it also includes sites and taxalist information. For each join, include an explicit argument saying which variable you are joining by (even if it will just assume the correct one for you). Store the updated data frame as birds_left. Make sure to look at the output - is what it contains consistent with what you expected it to contain?

birds_subset <- left_join(birds_subset, sites) |> 
  left_join(taxalist)
```

### full join 

```{r}
birds_full <- full_join(birds_subset, sites, by = "site_id")
```

### Inner join 

```{r}
#i expect inner_join() between birds_subset and taxalist column to have only columns shared between them ... because only species ID is conserved between them, I expect this to just tack on the columns for scientific and common names for the bords already in the subset 

birds_inner <- inner_join(birds_subset, taxalist, by = "species_id")

#yuuup I was right 

#what if I used left join instead?

birds_left <- left_join(birds_subset, taxalist, by = "species_id")
  
#seems to be spitting out the same thing 
#does not make sense to me 
```

# task 2

-   Starting with your `birds` object, rename the `notes` column to `bird_obs_notes` (so this doesn\'t conflict with `notes` in the `surveys`dataset

```{r}
bird_observations <- rename(.data = bird_observations, bird_obs_notes = notes)
```

-   Then, create a subset that contains *all observations* in the `birds` dataset, joins the taxonomic, site and survey information to it, and is finally limited to only columns `survey_date`, `common_name`, `park_name`, and `bird_count`. You can decide the order that you want to create this in (e.g. limit the columns first, then join, or the other way around).

```{r}
bird_observations_2 <- bird_observations |> 
  left_join(bird_observations, taxalist, by = "species_id")
```
